---
title: "Most Valued Data Science Skills"
author: "Escarlet Gabriel Vicente"
date: "2025-10-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

In today’s data-driven world, *data science* has become one of the most in-demand and interdisciplinary career paths.
Professionals in this field are expected to combine technical, analytical, and communication skills to extract insights and drive decisions using data.

This project explores the question:

“Which are the most valued data science skills?”

Using a dataset from *Kaggle (Data Science Job Postings & Skills, 2024)*, we analyze more than *12,000 job listings* to identify the most frequently requested skills among employers in the data science job market.

## Data Source
**Dataset:** [Data Science Job Postings & Skills (2024)](https://www.kaggle.com/datasets/asaniczka/data-science-job-postings-and-skills)  
**Author:** asaniczka  
**Platform:** Kaggle  
**Source:** LinkedIn job postings  
**License:** ODC Attribution License  

This dataset provides a raw dump of data science related job postings collected from LinkedIn.
It includes details such as job titles, companies, locations, and most importantly, a list of skills mentioned for each posting.
The main objective of the dataset is to allow users to practice data cleaning and to explore which skills are most relevant in the current job market.

```{r}
library(tidyverse)
```

## Step 1 - Import the Dataset
```{r}
job_skills <- read_csv("job_skills.csv")
```

## Step 2 - Explore the Data

```{r}
head(job_skills)
glimpse(job_skills)
```

## Step 3 - Clean and Transform the Skills Column
```{r}
skills_clean <- job_skills %>%
  separate_rows(job_skills, sep = ",") %>%        
  mutate(job_skills = str_trim(job_skills)) %>%   
  filter(job_skills != "")

skills_clean
```

## Step 4 - Count Skill Frequencies
```{r}
skill_counts <- skills_clean %>%
  count(job_skills, sort = TRUE)

head(skill_counts, 10)
```

## Step 5 - Top 10 Skills Visualization
```{r}
skill_counts %>%
  slice_max(n, n = 10) %>%
  ggplot(aes(x = reorder(job_skills, n), y = n)) +
  geom_col(fill = "orange") +
  coord_flip() +
  labs(
    title = "Top 10 Most Valued Data Science Skills (Job Postings)",
    x = "Skill",
    y = "Frequency"
  ) +
  theme_minimal()
```

## Summary
The analysis shows that Python, SQL, and Machine Learning are among the most frequently mentioned skills in job postings.
Communication, Data Analysis, and Visualization tools like Tableau and R also appear prominently, suggesting that data scientists must balance technical proficiency with analytical and storytelling skills.


