---
title: "Tidying and Transforming Country GDP and Population Data"
author: "Escarlet Gabriel Vicente"
date: "2025-10-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## 1. Introduction
In this part, I worked with a dataset showing the population and GDP for three countries (USA, China, and India) across three different years: 2000, 2005, and 2010.
The dataset is untidy because it is presented in a wide format, where each year has its own set of columns for both population and GDP.
The goal was to tidy the dataset, reshape it into a long format, and make it easier to analyze and visualize economic and population trends over time.

---

## 2. Load Libraries
To start, I loaded the libraries used for organizing, transforming, and analyzing the data.
```{r libraries}
library(tidyverse)
```

## 3. Create the Dataset
I recreated the dataset in wide format to match the structure of the original table.
This table contains population and GDP data for the years 2000, 2005, and 2010, stored in separate columns for each year.
```{r create_dataset}
gdp_data <- data.frame(
  Country = c("USA", "China", "India"),
  `2000_Population` = c(282162411, 1262645000, 1053050912),
  `2000_GDP` = c(10285, 1198, 476),
  `2005_Population` = c(295516599, 1307560000, 1139964932),
  `2005_GDP` = c(13094, 2286, 834),
  `2010_Population` = c(309327143, 1340910000, 1224614327),
  `2010_GDP` = c(14964, 6087, 1708)
)

gdp_data
```

## 4. Tidy the Data
The dataset was untidy because each year had separate columns for population and GDP.
To tidy it, I converted the dataset into long format using pivot_longer(), creating columns for year, population, and GDP.
This format makes it easier to perform comparisons and trend analyses across countries and time periods.
```{r tidy_data}
gdp_long <- gdp_data %>%
  pivot_longer(
    cols = -Country,
    names_to = c("Year", ".value"),
    names_sep = "_"
  )

gdp_long
```

## 5. Transform and Summary
Next, I summarized population and GDP changes over time to highlight growth patterns for each country.
This allowed me to observe how both variables evolved between 2000 and 2010.
```{r transformation}
summary_table <- gdp_long %>%
  group_by(Country) %>%
  summarise(
    Population_Growth = max(Population) - min(Population),
    GDP_Growth = max(GDP) - min(GDP)
  )

summary_table
```

## 6. Visualization
To visualize the economic trends, I created a line chart showing GDP changes over time for each country.
This chart illustrates how each nation’s economy developed during the decade, with China showing the fastest rate of GDP growth.
```{r line_chart}
ggplot(gdp_long, aes(x = Year, y = GDP, color = Country, group = Country)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(title = "GDP Growth (2000–2010)", y = "GDP per Capita ($)", x = "Year") +
  theme_minimal()
```

## 7. Conclusion
After tidying and transforming the dataset, I was able to clearly visualize population and GDP trends from 2000 to 2010.
This part of the project demonstrated how reshaping wide-format data into long format provides flexibility for comparison, statistical summaries, and visual analysis of growth over time.